---
title: "Module 6 Rtools: tidymodels revisited"
---

## Tidying up our models

The R supplement from Module 4 introduced the `tidymodels` package as a one-stop shop for fitting multiple models to training data and comparing resampling errors or test set performance. Here we'll review the details by fitting models from Module 6 using this alternative framework. 

Since I am sure we are all getting a bit tired of the election dataset, we'll introduce a newer and bigger one for use in this section: a database of Spotify songs with popularity, genre, and various other musical characteristics. First we'll load it and look at the variables:

```{r,message=F,warning=F}
library(tidyverse)
library(tidymodels)
library(broom)
library(rattle)
df_url="https://raw.githubusercontent.com/jmayberr/math133-stat-learning-fall2026/refs/heads/main/Datasets/spotify.csv"
spotify=read.csv(df_url) |> 
  drop_na() |>
  select(-track_id,-X) |>
  mutate(track_genre=as.factor(track_genre))
glimpse(spotify)
```

We'll demonstrate the use of `tidymodels` for regression scenarios using popularity as our response and all the other variables (minus artist, album, and track name) as predictors. While the codes below will work for the full dataset, they are very slow with that many observations so we will also apply them to a subset of observations from a few select genres^[Sorry, author bias.]

```{r}
genres=c("black-metal","death-metal","emo",
         "grindcore","grunge","hardcore",
         "heavy-metal","punk","punk-rock","metalcore")
spotify_small=spotify |>
  dplyr::filter(track_genre %in% genres)
```

Just for something new, here is how you can perform a training-testing split using tidymodels functions.

```{r}
set.seed(9)
s=initial_split(spotify_small,prop=0.7)
train_set=training(s)
test_set=testing(s)
```

## Regression Tree.

Its been a while (even for me) so let's work through the steps of building a tidymodel again, this time for a regression tree:

```{r}
# Step 1: Model Specification
tree_spec = decision_tree(
  tree_depth = 5,
  min_n=20,
  cost_complexity = 0.01
) |>
  set_engine("rpart") |>
  set_mode("regression")
```

A few notes here:

* Unlike `rpart` which chooses default values, this formulation requires you make a choice about three parameters in the tree: `tree_depth` (the maximum allowable tree depth),  `min_n` (the minimum number of data points required for a node to be split), and `cost_complexity` (the cp paramater from rpart). For now, I chose values above which roughly match the defaults (except for a smaller `tree_depth`). 
* You can hypertune based on these three parameters. We won't do that in these notes, but see the next section where we exemplify the process for random forests. 
* There are other engines available besides `rpart` that differ in their execution of the fitting algorithm; `spark` is a popular one. 
* For classification problems, obviously change the last line to "classification"

Now we move onto 
```{r}
# Step 2: Define resampling plan
cv_plan = vfold_cv(train_set[,4:19],v=10)

#Step 3: Evaluate model performance
tree_model=fit_resamples(tree_spec,
                       popularity~.,
                       resamples=cv_plan)

collect_metrics(tree_model)
```

Finally, if we're happy with this (we don't have too much of a benchmark but $R^2 \approx 0.3$ seems decent), we refit the model to the whole training set and plot the resulting tree

```{r}
# Step 4: Fit final model (and plot in this case)
tree_fit=fit(tree_spec,
             popularity~.,
             train_set[,4:19])
fancyRpartPlot(extract_fit_engine(tree_fit))
```
Its not too surprising that grindcore is at the bottom. Its a little more surprising that metalcore is up there with grunge and emo in popularity, but I don't mind that. Interestingly, energy plays an important role in further distinguishing songs from these top three genres.  

To evaluate performance on the test set, we can use our old friend predict and then calculate whatever metrics we want (eg. $R^2$). 

```{r}
tree_pred=predict(tree_fit,test_set)
SSE=sum((test_set$popularity-tree_pred)^2)
SST=sum((test_set$popularity-mean(test_set$popularity))^2)
1-SSE/SST
```

We could also tune the parameters to see if, for example, decreasing min_n might create additional splits and improve the fit but instead, let's move onto random forests and demonstrate that process there.  

## Hypertuning random forests

One advantage of using tidymodels for random forests is that we can tune the "mtry" hyperparameter which is hard to choose from scratch. Let's review that process step by step:

```{r}
# Step 1: Model specification
rf_specs=rand_forest(mtry=tune()) |>
  set_engine("ranger") |>  # can also use randomForest but ranger is usually preferred
  set_mode("regression")

# Step 2: Resampling procedure - switch to five folds for faster performance
cv_plan =vfold_cv(train_set[,4:19],v=5)

# Step 3: Parameter tuning/ Model evaluation

# Set mtry_grid
mtry_grid=grid_regular(
  finalize(mtry(),train_set[,5:19]), # This line gives mtry the range of 1 to max number of features
  levels = 5)  #Sets how many values in the range to try, can do 14 but its slower.  

rf_models=tune_grid(rf_specs,
                    popularity~.,
                    cv_plan,
                    grid=mtry_grid)
```

Then we can plot and choose the best

```{r}
autoplot(rf_models)
best_mtry=select_best(rf_models,metric="rmse")
cat("The best value of mtry is", best_mtry$mtry)
```

From there we fit the final model

```{r}
rf_best=finalize_model(rf_specs,best_mtry)
rf_fit=fit(rf_best,
           popularity~.,
           train_set[,4:19])
```

and evaluate predictive value. 

```{r}
rf_pred=predict(rf_fit,test_set)
SSE=sum((test_set$popularity-rf_pred)^2)
1-SSE/SST
```
That was a lot better than the tree model! 

One note on variable importance... 

# ```{r}
# rf_spec <- rand_forest(mtry = best_mtry) |>
#   set_engine("ranger", importance = "impurity") |>
#   set_mode("regression")  
# 
# ```