---
title: "Module 4 R Tools: Hyperparameter Tuning with Tidymodels"
format: html
---

## One Ring to Rule Them All 

Module 4 has primarily focused on alternative methods for finding coefficients in the linear regression model
$$
Y = X \beta + \varepsilon
$$
with a brief address of classification models as well. We have seen several different functions useful for estimating $\beta$:

* `lm()`: core R function which uses OLS to estimate $\beta$
* `regsubsets()`: function in the `leaps` package which determines the best OLS model with $i$ predictors for $i=1,...,p$ and produces complexity metrics for comparison
* `glmnet()`: function from the `glmnet` package which applies ridge, lasso, or elastic net to estimate $\beta$

We also saw that cross-validation is an important tool for model evaluation and hyper-parameter tuning. In Module 3, we developed KNN and logistic regression for classification problems and had to write our own loops to tune $K$ in KNN.

As our toolbox grows, it would be nice to have a unified framework that smooths out syntax across different models and automates the workflow hyper-parameter tuning. The `tidymodels` package is designed with exactly that purpose in mind^[Although like the one ring, it takes a toll on its user.]. In these notes we'll focus on its most powerful application: **hyperparameter tuning**.

## The Tidymodels Workflow

There is a four step process to using the `tidymodels` framework:

1. **Model specification:** specify which model to use and which R package ("engine") to use for fitting
2. **Resampling plan:** describe how to split the data for model evaluation (e.g. 10-fold cross-validation, LOO)
3. **Tuning and evaluation:** fit the specified model across a grid of hyperparameter values using the resampling plan and calculate model performance metrics (e.g. RMSE, $R^2$, accuracy)
4. **Final fit:** select the best hyperparameters and fit the final model to the full training dataset

To build up to hyperparameter tuning, let's first see how the workflow looks for a simple OLS model with no hyperparameters to tune.

## Warm Up: OLS with tidymodels

```{r}
# Step 0: Load data and packages
library(tidymodels)
set.seed(516)
df_url = "https://raw.githubusercontent.com/jmayberr/math133-stat-learning-fall2026/refs/heads/main/Datasets/election2016.csv"
election2016 = read.csv(df_url)

# Step 1: Model specification
lm_spec = linear_reg() |>
  set_engine("lm")

# Step 2: Resampling plan; using 10-Fold CV with 10 repeats for additional stability
cv_plan = vfold_cv(election2016[,2:10], v = 10, repeats = 10)

# Step 3: Evaluate (no hyperparameters to tune here)
lm_fit = fit_resamples(
  lm_spec,
  trump.vote ~ .,
  resamples = cv_plan
)

collect_metrics(lm_fit)
```

The `mean` column shows the average metric across all folds and `std_err` is its standard deviation across folds. Using `repeats = 10` gives a more stable estimate of the cross-validated RMSE than a single fold split. 

For OLS there are no hyperparameters so Step 3 uses `fit_resamples()` rather than `tune_grid()`. Once we are happy with the evaluation, Step 4 fits the final model to the full dataset:

```{r}
# Step 4: Fit final model
library(broom)
final_fit_lm = fit(lm_spec, trump.vote ~ ., data = election2016[,2:10])
tidy(final_fit_lm)
```

Note that if you don't care about cross-validation, you could jump straight to the final model with

```{r}
fit_lm= linear_reg() |>
  set_engine("lm") |>
  fit(trump.vote ~ ., 
      data = election2016[,2:10])
tidy(fit_lm)
```

This is fine but honestly `lm()` directly is simpler for OLS — the real payoff of tidymodels comes when we need to tune hyperparameters, as we'll see next.

## The Main Event: Hyperparameter Tuning

Now let's see where tidymodels really earns its keep. We'll tune the penalty parameter $\lambda$ in a ridge model, something that required the somewhat clunky `cv.glmnet()` workflow in the last notes. Note that "mixture" is the alpha from Module 4.3 and "penalty" is the lambda.     

```{r}
# Step 1: Model specification with penalty to tune
ridge_spec = linear_reg(penalty = tune(), mixture = 0) |>
  set_engine("glmnet")

# Step 2: Resampling plan; you can do repeats = 10 as well, but it takes longer to run 
cv_plan = vfold_cv(election2016[,2:10], v = 10)

# Step 3: Define grid of lambda values and tune
lambda_grid = grid_regular(penalty(), levels = 50)

ridge_fit = tune_grid(
  ridge_spec,
  trump.vote ~ .,
  resamples = cv_plan,
  grid = lambda_grid
)

# Visualize tuning results
autoplot(ridge_fit)

# Find best lambda
best_lambda = select_best(ridge_fit, metric = "rmse")
best_lambda
```

The `autoplot()` shows how RMSE changes across the grid of $\lambda$ values. You can see it decrease and then increase again as $\lambda$ gets too large, which is the bias-variance tradeoff in action!

After completing tuning, we can look at our final model. 

```{r}
# Step 4: Finalize and fit
final_ridge <- finalize_model(ridge_spec, best_lambda)
final_fit_ridge <- fit(final_ridge, trump.vote ~ ., data = election2016[,2:10])
tidy(final_fit_ridge)
```

Notice how clean this is compared to the `cv.glmnet()` workflow from the last notes. There is no manual matrix conversion, no `lambda.min` extraction, and `tidy()` gives us a clean coefficient table directly! 

## Tuning Both Parameters: Elastic Net

In the ridge example above we fixed `mixture = 0` and only tuned `penalty`. But we can also tune `mixture` alongside `penalty` to let the data decide whether ridge, lasso, or something in between works best. As we mentioned in Module 4.3, this is called **elastic net** and requires only a small change to the model specification:

```{r}
# Step 1: Elastic net with both parameters to tune
elastic_spec = linear_reg(penalty = tune(), mixture = tune()) |>
  set_engine("glmnet")

# Step 2: Resampling plan (same as before)
cv_plan = vfold_cv(election2016[,2:10], v = 10)

# Step 3: 10x10 grid over both parameters (100 models total) 
# Could do 20x20  by switching levels = 20 but it takes a while to run
elastic_grid = grid_regular(
  penalty(),
  mixture(),
  levels = 10
)

elastic_fit = tune_grid(
  elastic_spec,
  trump.vote ~ .,
  resamples = cv_plan,
  grid = elastic_grid
)

autoplot(elastic_fit)
best_params = select_best(elastic_fit, metric = "rmse")
best_params
```

The `autoplot()` is particularly informative here since it shows performance across both dimensions simultaneously. You can see at a glance whether ridge ($\alpha = 0$), lasso ($\alpha = 1$), or something in between works best for this data. In this case, it turns out ridge was best

```{r}
# Step 4: Finalize and fit
final_elastic <- finalize_model(elastic_spec, best_params)
final_fit_elastic <- fit(final_elastic, trump.vote ~ ., data = election2016[,2:10])
tidy(final_fit_elastic)
```

Notice that the only change from the ridge example was adding `mixture = tune()` to the model specification and including `mixture()` in the grid. Everything else (the resampling plan, `tune_grid()`, `select_best()`, and `finalize_model()`) stayed exactly the same. That's the one ring in action!

## Tuning K in KNN: The Same Workflow

As another example of where the "one ring" idea really pays off, recall from Module 3 that we had to write a clunky for loop to find the optimal $K$ in KNN. With tidymodels, the workflow is almost identical to the ridge example above. We just change the model specification!

We'll demonstrate by predicting the binary outcome in election2016. Note that you'll need to install and load the `kknn` package first. Note that we will decrease the number of folds to 5 here because with $K=10$ there are only 5 observations per fold and the accuracy will always be 0, 0.2, 0.4, 0.6, 0.8 or 1. Note also the addition of the "strata" argument for resampling. This ensures that when the folds are formed they will have representation from both classes of our outcome (red/blue). 

```{r,warning=F,message=F}
library(kknn)
# Step 1: KNN specification with K to tune
knn_spec = nearest_neighbor(neighbors = tune()) |>
  set_engine("kknn") |>
  set_mode("classification") # Default is regression 

# Step 2: Resampling plan (stratified by outcome for classification)
election2016$outcome=as.factor(election2016$outcome)
cv_plan = vfold_cv(election2016[,c(2:9,11)], v = 5,repeats=10,strata=outcome)

# Step 3: Tune K
k_grid = grid_regular(neighbors(range = c(1, 30)), levels = 30)

knn_fit = tune_grid(
  knn_spec,
  outcome ~ .,
  resamples = cv_plan,
  grid = k_grid,
  metrics = metric_set(accuracy, bal_accuracy)
)

autoplot(knn_fit)
best_k = select_best(knn_fit, metric = "accuracy")
best_k
```

```{r}
# Step 4: Finalize and fit
final_knn = finalize_model(knn_spec, best_k)
final_fit_knn = fit(final_knn, outcome ~ ., data = election2016[,c(2:9,11)])

```

Compare how much cleaner this is than the for loop from Module 3! And critically, the workflow is **identical** to the ridge example — only the model specification changed. That's the whole point of tidymodels.

## Logistic Regression with Elastic Net Tuning

You can also apply elastic net regularization to logistic regression which is useful when you have many predictors and want to do feature selection and classification simultaneously. The workflow is identical to the ridge/lasso examples after a couple small modifications to Step 1: 

```{r}
election2016$outcome = as.factor(election2016$outcome)

# Step 1: Logistic regression with elastic net penalties to tune
log_spec = logistic_reg(penalty = tune(), mixture = tune()) |>
  set_engine("glmnet") |>
  set_mode("classification")

# Step 2: Resampling plan (stratified by outcome since classes may be imbalanced)
cv_plan = vfold_cv(election2016[,c(2:9,11)], v = 5, repeats = 10,
                    strata = outcome)

# Step 3: Tune penalty and mixture
elastic_grid = grid_regular(
  penalty(),
  mixture(),
  levels = 10
)

log_fit = tune_grid(
  log_spec,
  outcome ~ .,
  resamples = cv_plan,
  grid = elastic_grid,
  metrics = metric_set(accuracy, bal_accuracy)
)

autoplot(log_fit)
best_params = select_best(log_fit, metric = "bal_accuracy")
best_params
# Step 4: Finalize and fit
final_log <- finalize_model(log_spec, best_params)
final_fit_log <- fit(final_log, outcome ~ ., data = election2016[,c(2:9,11)])
tidy(final_fit_log)
```





## Summary

For the three classes of models discussed in this section, the tidymodels workflow is:

| Step | No hyperparameters | With hyperparameters |
|------|-------------------|---------------------|
| **Spec** | `linear_reg() |> set_engine("lm")` | `linear_reg(penalty = tune()) |> set_engine("glmnet")` |
| | `logistic_reg() |> set_engine("glm")` | `logistic_reg(penalty = tune(), mixture = tune()) |> set_engine("glmnet")` |
| | `nearest_neighbor(neighbors = 5) |> set_engine("kknn")` | `nearest_neighbor(neighbors = tune()) |> set_engine("kknn")` |
| **Resample** | `vfold_cv(data, v = 10)` | `vfold_cv(data, v = 10)` |
| | `vfold_cv(data, v = 10, strata = outcome)` | `vfold_cv(data, v = 10, strata = outcome)` |
| **Evaluate** | `fit_resamples()` | `tune_grid(..., grid = my_grid)` |
| **Metrics** | `metric_set(rmse, rsq)` | `metric_set(rmse, rsq)` |
| | `metric_set(accuracy, bal_accuracy)` | `metric_set(accuracy, bal_accuracy)` |
| **Finalize** | `fit(spec, formula, data)` | `finalize_model(spec, best_params)` then `fit()` |
| **Extract** | `tidy(final_fit)` | `tidy(final_fit)` |

The key insight is that the workflow is identical regardless of whether you are tuning $\lambda$ in ridge, $K$ in KNN, or any other hyperparameter in any other model. That uniformity is what makes tidymodels worth learning!

## Practice

Repeat the practice problems from 4.3 using the `tidymodels` workflow (applied to the training data). Which do you prefer?

Bonus: Tune the model for both penalty and mixture using `tidymodels`. Compute the test RMSE for this model and compare with the optimal lasso/ridge from 4.3. 
