---
title: "Module 2 Theory"
---

These notes require a solid understanding of linear algebra at the level of an introductory course and probability at the level of a rigorous undergraduate course.

## The multivariate normal

Suppose that $Z$ is a vector of $n$ independent and identically distributed standard $N(0,1)$ random variables. Then any vector of the form $Y=\mu + AZ$ where $A$ is an $m\times n$ matrix and $\mu$ is a $m \times 1$ vector is said to have a multivariate normal (mvn) distribution with mean $\mu$ and variance $\Sigma = AA^T$. It can be shown that the joint density of $Y_1,...,Y_n$ is $$
f(y) = (2\pi)^{-m/2} |\det \Sigma|^{-1/2} e^{-\frac{1}{2}(y-\mu)^T \Sigma^{-1} (y-\mu)}.
$$Note that Fisher's Lemma says that if $Y$ is mvn with variance $\Sigma$ and $B$ is a matrix of appropriate dimensions, then $BY$ will also be mvn with variance $B \Sigma B^T$. We'll see the relevance of this below.

## Normal Equations

Suppose that $Y = X \beta + \varepsilon$ where $X$ is an $n \times p$ model matrix and let $$
|x|_2 = \sqrt{ \sum_{i=1}^n x_i^2} 
$$ denote the $\ell_2$-norm in $\mathbb{R}^n$. Then the least squares solution is the one that minimizes $$
|Y - X\beta|_2^2
$$ This is the same as the following problem from linear algebra: find the vector $\beta$ in the column space of $X$ which is as close as possible to $Y$. Such a vector $\hat y$ must satisfy $$
Y - \hat y \perp X_i
$$ for all columns $X_i$ of $X$ or in other words, $$
X^T (Y - X\beta) = 0
$$ Simplifying leads to the *normal equations*[^1] $$
X^TX \beta = X^T Y
$$ When $p < n$ (less predictors than observations), this system typically has a unique solution[^2] $$
\hat{\beta} = (X^TX)^{-1} X^T Y
$$ 
and then the predictions are 
$$
\hat{Y} = X \hat{\beta} = X(X^TX)^{-1} X^T Y = H Y
$$

[^1]: No relationship to normal random variables. Every subfield of math has its own ideas of what's normal and this is what is normal for numerical linear algebraists I guess

[^2]: Technically, we need $X$ to have full column rank.

where $H$ is that nasty looking matrix in front of $Y$ that can be easily formed in any software package. It often goes by the name *hat* (since it puts the "hat" on $Y$) or *projection* (since it projects onto the column space of $X$) matrix.

For example, to explicitly calculate the coefficients in R using this formula, we can use the `model.matrix` command to construct the model matrix $X$ from the model formula and automatically add the column of ones:

```{r}
library(tidyverse)
url="https://raw.githubusercontent.com/jmayberr/math133-stat-learning-fall2026/refs/heads/main/Datasets/Advertising.csv"
advertising=read_csv(url)
sales_formula=sales~TV+radio+newspaper
y=advertising$sales
X=model.matrix(sales_formula,data=advertising) 
head(X)

```

The matrix $X^T X$ can be formed by doing matrix multiplication which is a bit tedious in R, but I guess that is the drawback of making elementwise multiplication so easy:

```{r}
t(X) %*% X
```

Notice that this matrix is only $(p+1) \times (p+1)$ where $p=3$ is the number of predictors so it isn't computational expensive to invert. Thus we could calculate the least squares coefficients by

```{r}
betahat = solve(t(X) %*% X) %*% t(X) %*% y
print(betahat)
```

The `solve` command calculates the inverse. These are the same estimates we get from `lm`.

```{r}
lm(sales_formula,advertising)
```

If $p$ happens to be large, it is better to solve the system \$(X\^TX) \beta = X\^TY \$ directly using Gaussian elimination or other numeric techniques for solving systems of equations. This can also be done with `solve`:

```{r}
solve(t(X) %*% X,t(X) %*% y)
```

## Standard Errors and Coefficient Tests

Deriving properties of $\hat \beta$  is easier if we work in matrix form, for example: 
$$
E( \hat \beta) = (X^TX)^{-1} X^T E(Y) = (X^TX)^{-1} X^T X \beta = \beta
$$
Also since the covariance matrix of $Y$ is a diagonal matrix of $\sigma^2$ terms, 
$$
\text{var}( \hat \beta) = (X^TX)^{-1} X^T(\sigma^2 I) [(X^TX)^{-1} X^T]^T = \sigma^2 (X^TX)^{-1}.
$$

One can use the variances above to do individual t-tests for non-zero coefficients. Since 
$$
\hat \beta = (X^TX)^{-1} X^T Y
$$
and $Y$ is mvn with variance $\sigma^2I$, $\hat \beta$ is also mvn by Fisher's Lemma and its variance matrix is $\sigma^2 (X^TX)^{-1}$ as above. This implies that the variance of $\hat \beta_j$ is the $(j,j)$ entry of the matrix $\sigma^2 (X^TX)^{-1}$. Letting $C =(X^TX)^{-1}$, this implies that under the null hypothesis $\beta_j = 0$, $$
Z = \frac{\hat \beta_j}{\sigma \sqrt{c_{jj}}}
$$ is standard normal where $c_{jj}$ is the $(j,j)$ entry of $C$. Replacing $\sigma$ with the residual standard deviation 
$$
\sigma_r = \sqrt{\frac{1}{n-p-1} \sum(y_i - \hat y_i)^2}
$$

gives a $t$-statistic with $n-p-1$ degrees of freedom. These are the t-statistics report by `summary(lm(...))` and the denominator is the corresponding standard error!

To see this explicitly for the variable $TV$ in the advertising dataset:

```{r}
sales_formula=sales~TV+radio+newspaper
y=advertising$sales
X=model.matrix(sales_formula,data=advertising) 
C=solve(t(X) %*% X)
betahat= C %*% t(X) %*% y
yhat=X %*% betahat
n=nrow(advertising)
p=3
sigma_r= sqrt(sum((y-yhat)^2)/(n-p-1)) 
se=sigma_r*sqrt(C[2,2])
t=betahat[2]/se
cat("t =",t, "\nse =",se)
```

```{r}
summary(lm(sales_formula,advertising))
```

## Leverage and the American Dream

The leverage of the $i$th point in a dataset is defined as the rate of change of the prediction $\hat{y}_i$ as we change the value of $y_i$:
$$
\frac{\partial \hat{y}_i}{\partial y_i}
$$
Recalling that 
$$
\hat{Y} = H Y
$$
where $H = X(X^TX)^{-1} X^T$, we have 
$$
\frac{\partial \hat{y}_i}{\partial y_i} = H_{ii}
$$ 
Hence, the leverages are just the diagonal entries of the hat matrix:

```{r}
h=diag(X %*% solve(t(X) %*% X) %*% t(X))
head(h)
```


